Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-06-04 Manuscript/manuscript.Rmd
specified tuning in methods

### FEATURE REDUCTION

@@

In order to suitably reduce the feature space while also maintaining the inclusion of possibly colocalized annotations, we implemented an elastic-net regularization technique prior to modelling. The technique using a combination of the *l_{1}* and *l_{2}* penalties. For each elastic-net regularization, 10-fold cross-validation was used to tune the parameters that controlled the penalty terms. <!--The parameters that maximized the mean AUPRC across each cross-fold was chosen.--> All elastic-net algorithms were implemented in R using the *caret* package version 6.0 [@kuhn2012caret].

@@

Once the appropriate predictor type and set of features were determined, a final set of random forest classification models were built to compare model performances between re-sampling techniques and predictor types. <!--To reduce bias due to random dataset generation, 10-fold cross validation was used to tune the number of features to consider at each node and the number of ensemble trees to aggregate.--> The default number of features to consider at each node of the random forest algorithm was set as $\sqrt{p}$, where *p* refers to the number of predictors included in the model. Likewise, the number of ensemble trees to aggregate was set at 500. All random forest algorithms were implemented in R using the *caret* package version 6.0 [@kuhn2012caret]. Once the model was implemented, it was validated on the test set.

@@

To systematically summarize and compare our results of data driven random forest models, we created an ensemble framework (Figure 3). Each system in the framework was assumed to be composed of a model constructed from a particular cell line, $L = \{GM12878, K562\}$. Within each cell line we evaluated 4 data resolutions, $R = \{10 kb, 25 kb, 50 kb, 100 kb\}$. We then considered 3 separate predictor types as the feature space including $P = \{OC, OP, Distance\}$. To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling, including $B = \{None, RUS, ROS, SMOTE\}$. Each combination of the above parameters was implemented on chromosome specific sections of the linear genome. Performances were then aggregated across each of the 21 chromosomes (omitting chromosome 9). Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling 2016 ($2 \times 4 \times 3 \times 4 \times 21$) separate models. We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials.

@@

We then went on the evaluate the effect of class imbalance on model performance. In order to set a baseline in performance, we first evaluated the least granular model, that is one implemented using OC predictor types, without data re-sampling. We found that models built on data with fully imbalanced classes exhibited poor MCCs across all resolutions (Figure 4A). We gradually increased the granularity by assessing models built on OP predictor types, while still not performing any data resampling. Here, the results were similar to that of OC types (Figure 4B). However, when including distance-type predictors, models were found to exhibit a marked improvement in MCC compared to the other two types (Figure 4C). Results were found to be consistently poor across resolutions and predictor types, when implemented on the K562 cell line (Supplementary Figure 3). Furthermore, we found that conventional metrics of performance (i.e. accuracy, sensitivity and specificity, and AUC) were less sensitive for models built on imbalanced data, regardless of cell line, resolution, or predictor-type (Supplementary Tables 6-7). [??? Need to know why PRCs give different results.]

>>>>

<!--
### FEATURE REDUCTION

@@

In order to suitably reduce the feature space while also maintaining the inclusion of possibly colocalized annotations, we implemented an elastic-net regularization technique prior to modelling. The technique using a combination of the *l_{1}* and *l_{2}* penalties. For each elastic-net regularization, 10-fold cross-validation was used to tune the parameters that controlled the penalty terms. <!--The parameters that maximized the mean AUPRC across each cross-fold was chosen.--> All elastic-net algorithms were implemented in R using the *caret* package version 6.0 [@kuhn2012caret]. 
-->

@@

Once the appropriate predictor type and set of features were determined, a final set of random forest classification models were built to compare model performances between re-sampling techniques and predictor types. To reduce bias due to random dataset generation, 10-fold cross validation was used to tune the number of features, *p*, to consider at each node. The number of trees that were aggregated for each random forest model was set to the default, n=500. All random forest algorithms were implemented in R using the *caret* package version 6.0 [@kuhn2012caret]. Once the model was implemented, it was validated on the test set.

@@

We created an ensemble framework to systematically compare different data driven characteristics to identify the optimal set of parameters when predicting TAD boundaries using functional genomic annotations (Figure 3). Each system in the framework was assumed to be composed of a random forest model constructed from a particular cell line, $L = \{GM12878, K562\}$. Within each cell line we evaluated 4 data resolutions, $R = \{10 kb, 25 kb, 50 kb, 100 kb\}$. We then considered 3 separate predictor types as the feature space including $P = \{OC, OP, Distance\}$. To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling, including $B = \{None, RUS, ROS, SMOTE\}$. Each combination of the above parameters was implemented on chromosome specific sections of the linear genome. Performances were then aggregated across each of the 21 chromosomes (omitting chromosome 9). Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling 2016 ($2 \times 4 \times 3 \times 4 \times 21$) separate models. We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials.

@@

<!--We then went on the evaluate the effect of class imbalance on model performance.--> In order to set a baseline in performance, we first evaluated the least granular model, that is one implemented using OC predictor types, without data re-sampling. We found that models built on data with fully imbalanced classes exhibited poor MCCs across all resolutions (Figure 4A). We gradually increased the granularity by assessing models built on OP predictor types, while still not performing any data resampling. Here, the results were similar to that of OC types (Figure 4B). However, when including distance-type predictors, models were found to exhibit a marked improvement in MCC compared to the other two types (Figure 4C). Results were found to be consistently poor across resolutions and predictor types, when implemented on the K562 cell line (Supplementary Figure 3). Furthermore, we found that conventional metrics of performance (i.e. accuracy, sensitivity and specificity, and AUC) were less sensitive to models built on imbalanced data, regardless of cell line, resolution, or predictor-type (Supplementary Tables 6-7).
